cardinalR: Generating interesting high-dimensional data structures

### 2025-09-16 paper-cardinalR.Rmd
Added how clusters can be map with branching data

A branching structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch)"} else { "Figure \\@ref(fig:branch-proj)"})`) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometry, number of points, and the number of branches. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:branching-tb-html)', '\\@ref(tab:branching-tb-pdf)'))` outlines these functions. The main arguments of the functions described in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:arg-branching-tb-html)', '\\@ref(tab:arg-branching-tb-pdf)'))`.

@@

```{r}
arg_branching_tb <- tibble(arg = c("n",
                          "p",
                          "k"), 
                        exp = c("A numeric value representing the number of points.",
                                "A numeric value representing the number of dimensions.",
                                "A numeric value representing the number of branches."))
```

@@

Each branch $i$ is constructed using $X_1 \sim U(-2, 2)$ and $X_2 = \exp(\pm s_iX_1) + \epsilon$, where: $\epsilon \sim U(0, 0.1)$ adds local jitter, $s_i \in [0.5, 2]$ is randomly sampled for each branch. The sign alternates between branches (odd: negative, even: positive exponent). This creates mirror-symmetric branches with different steepness and curvature where odd-numbered branches represent a decay pattern (reflected) and even-numbered branches represent a growth pattern. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.1^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

@@

The `gen_linearbranches(n, p, k)` function generates a dataset of $n$ points in $p-D$ forming $k$ linear branches. Each branch is a line segment with added jitter to simulate noise. Branches $1$ and $2$ are initialized with fixed slopes and intercepts: branch $1$ has slope $0.5$ and intercept $0$, with $X_1 \sim U(-2, 8)$ and $X_2 = 0.5 X_1 + \epsilon$, $\epsilon \sim U(0, 0.5)$; branch $2$ has slope $-0.5$ and intercept $0$, with $X_1 \sim U(-6, 2)$ and $X_2 = -0.5 X_1 + \epsilon$, $\epsilon \sim U(0, 0.5)$.

Branches $3$ to $k$ are added iteratively. Each new branch starts at a location outside predefined exclusion zones to avoid overlap with the first two branches. Specifically, $X_1$ starting points are sampled outside the intervals $[-8, -7]$, $[-2, 2]$, and $[7, 8]$, and $X_2$ starting points are sampled outside $[7, 8]$. For each new branch $i$, $X_1$ values are drawn over a short interval $[x_{\text{start}}, x_{\text{start}} + 1]$, and $X_2$ values are calculated as $X_2 = s_i (X_1 - x_{\text{start}}) + y_{\text{start}} + \epsilon$, where $s_i$ is a randomly chosen slope from a filtered sequence and $\epsilon \sim U(0, 0.2)$.

For $p > 2$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, for $j = 3, \dots, p$.

@@

This function generates $n$ points in $p-D$, where the last dimension, $X_p$, represents the height along the cone's axis, and the first $p-1$ dimensions define a shrinking hyperspherical cross-section toward the tip. Heights are sampled from a truncated exponential distribution, $X_p \sim \text{Exp}(\lambda = 2/h)$, capped at the cone height $h$, producing a higher density of points near the tip. At each height $X_p$, the radius of the cross-section decreases linearly from base to tip according to $r = r_{\text{min}} + (r_{\text{max}} - r_{\text{min}}) X_p / h$, where $r_{\text{min}} = \text{ratio}$ and $r_{\text{max}} = 1$.

>>>>

A branching structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch)"} else { "Figure \\@ref(fig:branch-proj)"})`) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometries, numbers of points, and number of branches. Although these functions can generate multiple branches, they do not produce a formal *multicluster* dataset: the branches form a single connected structure, with multiple visually distinct arms rather than independent clusters. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:branching-tb-html)', '\\@ref(tab:branching-tb-pdf)'))` outlines these functions. The main arguments of the functions described in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:arg-branching-tb-html)', '\\@ref(tab:arg-branching-tb-pdf)'))`.

@@

```{r}
arg_branching_tb <- tibble(arg = c("n",
                          "p",
                          "k"), 
                        exp = c("A numeric value representing the number of points.",
                                "A numeric value representing the number of dimensions.",
                                "A numeric value representing the number of clusters."))
```

@@

Each branch $i$ is constructed using $X_1 \sim U(-2, 2)$ and $X_2 = \exp(\pm s_iX_1) + \epsilon$, where: $\epsilon \sim U(0, 0.1)$ adds local jitter, $s_i \in [0.5, 2]$ is randomly sampled for each branch. The sign of the exponent alternates between branches: odd-numbered branches have a negative exponent (decaying pattern), and even-numbered branches have a positive exponent (growing pattern), producing mirror-symmetric curves with varying steepness. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.1^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

@@

The `gen_linearbranches(n, p, k)` function generates a dataset of $n$ points in $p\text{-}D$ forming $k$ linear branches. Each branch is a line segment with added jitter to simulate noise. Branches $1$ and $2$ are initialized with fixed slopes and intercepts: branch $1$ has slope $0.5$ and intercept $0$, with $X_1 \sim U(-2, 8)$ and $X_2 = 0.5 X_1 + \epsilon$, $\epsilon \sim U(0, 0.5)$; branch $2$ has slope $-0.5$ and intercept $0$, with $X_1 \sim U(-6, 2)$ and $X_2 = -0.5 X_1 + \epsilon$, $\epsilon \sim U(0, 0.5)$.

Branches $3$ to $k$ are added iteratively. Each new branch starts at a location outside predefined exclusion zones to avoid overlap with the first two branches. Specifically, $X_1$ starting points are sampled outside the intervals $[-8, -7]$, $[-2, 2]$, and $[7, 8]$, and $X_2$ starting points are sampled outside $[7, 8]$. For each new branch $i$, $X_1$ values are drawn over a short interval $[x_{\text{start}}, x_{\text{start}} + 1]$, and $X_2$ values are calculated as $X_2 = s_i (X_1 - x_{\text{start}}) + y_{\text{start}} + \epsilon$, where $s_i$ is a randomly chosen slope from a filtered sequence and $\epsilon \sim U(0, 0.2)$. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, for $j = 3, \dots, p$.

@@

This function generates $n$ points in $p\text{-}D$, where the last dimension, $X_p$, represents the height along the cone's axis, and the first $p-1$ dimensions define a shrinking hyperspherical cross-section toward the tip. Heights are sampled from a truncated exponential distribution, $X_p \sim \text{Exp}(\lambda = 2/h)$, capped at the cone height $h$, producing a higher density of points near the tip. At each height $X_p$, the radius of the cross-section decreases linearly from base to tip according to $r = r_{\text{min}} + (r_{\text{max}} - r_{\text{min}}) X_p / h$, where $r_{\text{min}} = \text{ratio}$ and $r_{\text{max}} = 1$.

### 2025-09-16 paper-cardinalR.tex
Added how clusters can be map with branching data

A branching structure (Figure \ref{fig:branch-proj}) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometry, number of points, and the number of branches. Table \ref{tab:branching-tb-pdf} outlines these functions. The main arguments of the functions described in Table \ref{tab:arg-branching-tb-pdf}.

@@

\caption{\label{tab:arg-branching-tb-pdf}The main arguments for branching shape generators.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Argument & Explanation\\
\midrule
n & A numeric value representing the number of points.\\
p & A numeric value representing the number of dimensions.\\
k & A numeric value representing the number of branches.\\
\bottomrule
\end{tabular}
\end{table}

@@

Each branch \(i\) is constructed using \(X_1 \sim U(-2, 2)\) and \(X_2 = \exp(\pm s_iX_1) + \epsilon\), where: \(\epsilon \sim U(0, 0.1)\) adds local jitter, \(s_i \in [0.5, 2]\) is randomly sampled for each branch. The sign alternates between branches (odd: negative, even: positive exponent). This creates mirror-symmetric branches with different steepness and curvature where odd-numbered branches represent a decay pattern (reflected) and even-numbered branches represent a growth pattern. For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.1^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

The \texttt{gen\_linearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points in \(p-D\) forming \(k\) linear branches. Each branch is a line segment with added jitter to simulate noise. Branches \(1\) and \(2\) are initialized with fixed slopes and intercepts: branch \(1\) has slope \(0.5\) and intercept \(0\), with \(X_1 \sim U(-2, 8)\) and \(X_2 = 0.5 X_1 + \epsilon\), \(\epsilon \sim U(0, 0.5)\); branch \(2\) has slope \(-0.5\) and intercept \(0\), with \(X_1 \sim U(-6, 2)\) and \(X_2 = -0.5 X_1 + \epsilon\), \(\epsilon \sim U(0, 0.5)\).

Branches \(3\) to \(k\) are added iteratively. Each new branch starts at a location outside predefined exclusion zones to avoid overlap with the first two branches. Specifically, \(X_1\) starting points are sampled outside the intervals \([-8, -7]\), \([-2, 2]\), and \([7, 8]\), and \(X_2\) starting points are sampled outside \([7, 8]\). For each new branch \(i\), \(X_1\) values are drawn over a short interval \([x_{\text{start}}, x_{\text{start}} + 1]\), and \(X_2\) values are calculated as \(X_2 = s_i (X_1 - x_{\text{start}}) + y_{\text{start}} + \epsilon\), where \(s_i\) is a randomly chosen slope from a filtered sequence and \(\epsilon \sim U(0, 0.2)\).

For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), for \(j = 3, \dots, p\).

@@

This function generates \(n\) points in \(p-D\), where the last dimension, \(X_p\), represents the height along the cone's axis, and the first \(p-1\) dimensions define a shrinking hyperspherical cross-section toward the tip. Heights are sampled from a truncated exponential distribution, \(X_p \sim \text{Exp}(\lambda = 2/h)\), capped at the cone height \(h\), producing a higher density of points near the tip. At each height \(X_p\), the radius of the cross-section decreases linearly from base to tip according to \(r = r_{\text{min}} + (r_{\text{max}} - r_{\text{min}}) X_p / h\), where \(r_{\text{min}} = \text{ratio}\) and \(r_{\text{max}} = 1\).

>>>>

A branching structure (Figure \ref{fig:branch-proj}) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometries, numbers of points, and number of branches. Although these functions can generate multiple branches, they do not produce a formal \emph{multicluster} dataset: the branches form a single connected structure, with multiple visually distinct arms rather than independent clusters. Table \ref{tab:branching-tb-pdf} outlines these functions. The main arguments of the functions described in Table \ref{tab:arg-branching-tb-pdf}.

@@

\caption{\label{tab:arg-branching-tb-pdf}The main arguments for branching shape generators.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Argument & Explanation\\
\midrule
n & A numeric value representing the number of points.\\
p & A numeric value representing the number of dimensions.\\
k & A numeric value representing the number of clusters.\\
\bottomrule
\end{tabular}
\end{table}

@@

Each branch \(i\) is constructed using \(X_1 \sim U(-2, 2)\) and \(X_2 = \exp(\pm s_iX_1) + \epsilon\), where: \(\epsilon \sim U(0, 0.1)\) adds local jitter, \(s_i \in [0.5, 2]\) is randomly sampled for each branch. The sign of the exponent alternates between branches: odd-numbered branches have a negative exponent (decaying pattern), and even-numbered branches have a positive exponent (growing pattern), producing mirror-symmetric curves with varying steepness. For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.1^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

The \texttt{gen\_linearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points in \(p\text{-}D\) forming \(k\) linear branches. Each branch is a line segment with added jitter to simulate noise. Branches \(1\) and \(2\) are initialized with fixed slopes and intercepts: branch \(1\) has slope \(0.5\) and intercept \(0\), with \(X_1 \sim U(-2, 8)\) and \(X_2 = 0.5 X_1 + \epsilon\), \(\epsilon \sim U(0, 0.5)\); branch \(2\) has slope \(-0.5\) and intercept \(0\), with \(X_1 \sim U(-6, 2)\) and \(X_2 = -0.5 X_1 + \epsilon\), \(\epsilon \sim U(0, 0.5)\).

Branches \(3\) to \(k\) are added iteratively. Each new branch starts at a location outside predefined exclusion zones to avoid overlap with the first two branches. Specifically, \(X_1\) starting points are sampled outside the intervals \([-8, -7]\), \([-2, 2]\), and \([7, 8]\), and \(X_2\) starting points are sampled outside \([7, 8]\). For each new branch \(i\), \(X_1\) values are drawn over a short interval \([x_{\text{start}}, x_{\text{start}} + 1]\), and \(X_2\) values are calculated as \(X_2 = s_i (X_1 - x_{\text{start}}) + y_{\text{start}} + \epsilon\), where \(s_i\) is a randomly chosen slope from a filtered sequence and \(\epsilon \sim U(0, 0.2)\). For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), for \(j = 3, \dots, p\).

@@

This function generates \(n\) points in \(p\text{-}D\), where the last dimension, \(X_p\), represents the height along the cone's axis, and the first \(p-1\) dimensions define a shrinking hyperspherical cross-section toward the tip. Heights are sampled from a truncated exponential distribution, \(X_p \sim \text{Exp}(\lambda = 2/h)\), capped at the cone height \(h\), producing a higher density of points near the tip. At each height \(X_p\), the radius of the cross-section decreases linearly from base to tip according to \(r = r_{\text{min}} + (r_{\text{max}} - r_{\text{min}}) X_p / h\), where \(r_{\text{min}} = \text{ratio}\) and \(r_{\text{max}} = 1\).
